---
title: "Getting started with BioRosa"
author: "Joel Correa da Rosa"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Getting started with BioRosa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE          # most chunks are illustrative; see note below
)
```

`BioRosa` packages the gene-expression analysis toolkit used in the Guttman
laboratory: richly **annotated heatmaps**, a one-call **enrichment pipeline**
(FGSEA + ORA + GSEA + an OpenXGR-style Fisher engine), the lab's curated
**pathway list**, and the helpers that tie a differential-expression result
together. This vignette walks through the main workflow end to end.

> **A note on the code chunks.** Most chunks below are shown for illustration
> only (`eval = FALSE`): the heatmap and enrichment steps depend on
> Bioconductor packages, write files, and can take minutes. The lightweight
> sections — exploring the bundled pathway list and the OpenXGR-style
> enrichment engine — are actually run.

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("correadarosaj/BioRosa")
```

```{r load, eval = TRUE, message = FALSE, warning = FALSE}
library(BioRosa)
```

# The BigTab: BioRosa's central data structure

Most BioRosa functions revolve around a **BigTab** — a gene-indexed data frame
that holds, for *each contrast*, a set of statistics in consistently named
columns. For a contrast suffix such as `Drug.W16vsBL` the columns are:

| Column                              | Meaning                                  |
|-------------------------------------|------------------------------------------|
| `lgFCH_Drug.W16vsBL`                | log2 fold change                         |
| `FCH_Drug.W16vsBL`                  | signed linear fold change `sign·2^|lgFCH|` |
| `pvals_Drug.W16vsBL`                | p-value                                  |
| `fdrs_Drug.W16vsBL`                 | adjusted p-value (FDR)                   |
| `StatusFCH1.3FDR0.05_Drug.W16vsBL`  | DEG call: `1` = up, `-1` = down, `0` = n.s. |

Genes live in a `GENENAME` column (or the row names). Contrast suffixes are
free-form and dot-separated, which is what lets the heatmap and enrichment
functions parse groups out of names like `R.Drug.W2vsBL` or `Lesional.W4vsBL`.

## Building a BigTab from a model fit

If you fit per-gene linear / mixed models, `runEbfit()` reshapes a tidy table
of estimates into an `ebfit` (a list with `$coef` and `$p.value`), and
`runBigTab()` turns that into a BigTab with all the columns above:

```{r build-bigtab}
# `tb` is a long table with columns: biomarker, contrasts, estimate, p.value
ebfit  <- runEbfit(tb)
BigTab <- runBigTab(ebfit, adj = "BH", mcut = 1.3, pcut = 0.05)
# -> columns lgFCH_*, FCH_*, pvals_*, fdrs_*, StatusFCH1.3FDR0.05_*
```

Already have a DESeq2 / limma table? You can build the equivalent columns
yourself, or skip the BigTab and pass parallel vectors directly to the
enrichment pipeline (shown later).

## Summarising the differential expression

`process_degs()` counts up/down genes per contrast from the `Status_*` columns
and writes one Excel workbook (a `Summary` sheet plus one sheet of DEGs per
contrast):

```{r process-degs}
res <- process_degs(
  BigTab      = BigTab,
  cont.list   = c("Drug.W2vsBL", "Drug.W16vsBL"),
  FCH         = 1.3,
  pcut        = 0.05,
  method      = "fdr",
  output_xlsx = "DEG_summary.xlsx"
)
res$summary          # per-contrast Up / Down counts
res$degs$Drug.W16vsBL  # DEG symbols for one contrast
```

# Annotated heatmaps

`doAnnotatedHeatmapGrid()` draws a publication-quality heatmap: a row-Z-scored
expression body on the left, and a companion **grid table** on the right where
every cell carries the signed fold change + significance stars for one
contrast, decimal-aligned per column. Contrast groups (parsed from the contrast
names) become boxed, colour-filled header bands.

```{r heatmap}
doAnnotatedHeatmapGrid(
  grafname   = "heatmap.pdf",
  mat        = expr_means,            # features x sample/group-mean columns
  annot.col  = annotx,                # row names define the column order
  colsplitx  = annotx$response,       # split the body into column groups
  BigTab     = BigTab,                # FCH_* / pvals_* columns
  BigTabcols = c("R.Drug.W2vsBL", "NR.Drug.W2vsBL", "RvsNR.Drug.W2vsBL"),
  setname    = "Immune proteins")
```

The function auto-sizes the canvas and fonts to the gene-set size, parses the
group bands from the contrast names (here `R` / `NR` / `RvsNR`), and drops any
dot-token that is constant across contrasts from the sub-labels. Everything is
overridable — see `?doAnnotatedHeatmapGrid`. The classic side-table variant is
`doAnnotatedHeatmap()`.

# One-call enrichment: `enrichment_onestep()`

`enrichment_onestep()` runs the full pathway-analysis battery in a single call
and writes every table and figure to `output_dir`. It is built around an
important distinction:

* **GSEA** (`fgsea` and `gseGO`) runs **once on the full ranked gene list** —
  direction is read from the sign of the NES. (Running GSEA on a pre-filtered or
  up/down-split list is statistically wrong.)
* **ORA** (`enrichGO` / `enrichKEGG` / `enrichPathway` / Hallmark, plus the
  OpenXGR-style RXGR engine) runs on the **UP and DOWN gene sets** — where the
  split is valid.

## From a BigTab + contrast

The most convenient entry point reads the log fold changes and p-values
straight from a BigTab for one contrast suffix:

```{r enrich-bigtab}
res <- enrichment_onestep(
  BigTab     = BigTab,
  contrast   = "Drug.W16vsBL",   # reads lgFCH_Drug.W16vsBL + pvals_Drug.W16vsBL
  output_dir = "results/Drug.W16vsBL"
)
```

## From parallel vectors

```{r enrich-vectors}
res <- enrichment_onestep(
  genes          = deg$SYMBOL,
  log2FoldChange = deg$log2FoldChange,
  padj           = deg$padj,
  output_dir     = "results/enrichment"
)
```

## Useful options

```{r enrich-options}
res <- enrichment_onestep(
  BigTab   = BigTab,
  contrast = "Drug.W16vsBL",

  # 1. Define the ORA UP/DOWN sets from the BigTab status column (1 = up,
  #    -1 = down) instead of recomputing from thresholds. On by default when
  #    a status column is present.
  ora_from_status = TRUE,
  status_prefix   = "Status",     # matches Status_... and StatusFCH1.3FDR0.05_...

  # 2. Choose the GSEA ranking statistic. "FCH" amplifies large fold changes
  #    in the weighted enrichment score; "signed_pval" = sign(lgFCH)*-log10(p).
  rank_by = "lgFCH",              # or "FCH", "signed_pval"

  # 3. Toggle the OpenXGR-style Fisher / odds-ratio engine (on by default).
  run_rxgr = TRUE,

  output_dir = "results/Drug.W16vsBL"
)
```

The returned list has four elements:

* `res$fgsea` — per-collection FGSEA (`Hallmark`, `GO_BP`, `Reactome`).
* `res$ora`   — `$UP` / `$DOWN`, each with `GO`, `Reactome`, `KEGG`, `Hallmark`.
* `res$gsea`  — a single `gseGO` result.
* `res$rxgr`  — `$UP` / `$DOWN`, each a per-collection RXGR result
  (`$table`, `$dotplot`, `$forest`).

…and `output_dir/` is populated with the matching Excel tables and
dotplot / treeplot / cnetplot / ridgeplot / forest figures.

# The OpenXGR-style enrichment engine (RXGR)

The Fisher's-exact engine that powers `res$rxgr` is also exposed directly. It
takes a query gene vector and any **named list of gene sets** — including the
bundled `guttman_pathways` — and reports the same statistics as the OpenXGR web
server: enrichment z-score, odds ratio with 95% CI, and FDR.

```{r osea, eval = TRUE, message = FALSE, fig.width = 6, fig.height = 4}
# A small, self-contained example (no Bioconductor / network needed).
sets  <- guttman_pathways[lengths(guttman_pathways) > 0]
query <- sets[["Immune Genes (EG)"]]

eset <- oSEA(query, sets, min.overlap = 3, size.range = c(5, 3000),
             verbose = FALSE)
head(oSEAextract(eset, sortBy = "or")[, c("name", "nO", "or", "adjp")])
```

The two companion figures — a dotplot (z-score vs significance) and a forest
plot (log2 odds ratio with 95% CIs) — are produced from the same object:

```{r osea-figs, eval = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5}
oSEAforest(eset, top = 10)
```

`rxgr_enrichment()` is the one-call wrapper that runs `oSEA()` and writes the
table + both figures to disk, mirroring the OpenXGR downloads.

# The bundled pathway list

`guttman_pathways` is the lab's curated gene-set collection (May 2026 release):
a named list of gene-symbol vectors, with many signatures split into directional
`... Up` / `... Down` halves.

```{r data, eval = TRUE}
length(guttman_pathways)                       # number of gene sets
head(names(guttman_pathways), 4)               # set names
length(guttman_pathways[["Immune Genes (EG)"]])  # genes in one set
```

A common pattern is to restrict a gene set to what you actually measured before
feeding it to a heatmap or scoring routine:

```{r data-subset}
present <- intersect(guttman_pathways[["Immune Genes (EG)"]], rownames(expr_mat))
```

# Where to go next

* `?doAnnotatedHeatmapGrid` — annotated grid heatmaps, all styling arguments.
* `?enrichment_onestep` — the full enrichment pipeline and its options.
* `?oSEA`, `?oSEAforest`, `?oSEAdotplot` — the OpenXGR-style engine.
* `?process_degs`, `?runBigTab` — building and summarising a BigTab.

```{r session, eval = TRUE}
sessionInfo()
```
