---
title: "Getting started with BioRosa"
author: "Joel Correa da Rosa"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Getting started with BioRosa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE          # most chunks are illustrative; see note below
)
```

`BioRosa` packages the gene-expression analysis toolkit used in the Guttman
laboratory: richly **annotated heatmaps**, a one-call **enrichment pipeline**
(FGSEA + ORA + GSEA + an OpenXGR-style Fisher engine), the lab's curated
**pathway list**, and the helpers that tie a differential-expression result
together. This vignette walks through the main workflow end to end.

> **A note on the code chunks.** Most chunks below are shown for illustration
> only (`eval = FALSE`): the heatmap and enrichment steps depend on
> Bioconductor packages, write files, and can take minutes. The lightweight
> sections — exploring the bundled pathway list and the OpenXGR-style
> enrichment engine — are actually run.

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("correadarosaj/BioRosa")
```

```{r load, eval = TRUE, message = FALSE, warning = FALSE}
library(BioRosa)
```

# The BigTab: BioRosa's central data structure

Most BioRosa functions revolve around a **BigTab** — a gene-indexed data frame
that holds, for *each contrast*, a set of statistics in consistently named
columns. For a contrast suffix such as `Drug.W16vsBL` the columns are:

| Column                              | Meaning                                  |
|-------------------------------------|------------------------------------------|
| `lgFCH_Drug.W16vsBL`                | log2 fold change                         |
| `FCH_Drug.W16vsBL`                  | signed linear fold change `sign·2^|lgFCH|` |
| `pvals_Drug.W16vsBL`                | p-value                                  |
| `fdrs_Drug.W16vsBL`                 | adjusted p-value (FDR)                   |
| `StatusFCH1.3FDR0.05_Drug.W16vsBL`  | DEG call: `1` = up, `-1` = down, `0` = n.s. |

Genes live in a `GENENAME` column (or the row names). Contrast suffixes are
free-form and dot-separated, which is what lets the heatmap and enrichment
functions parse groups out of names like `R.Drug.W2vsBL` or `Lesional.W4vsBL`.

## Building a BigTab from a model fit

If you fit per-gene linear / mixed models, `runEbfit()` reshapes a tidy table
of estimates into an `ebfit` (a list with `$coef` and `$p.value`), and
`runBigTab()` turns that into a BigTab with all the columns above:

```{r build-bigtab}
# `tb` is a long table with columns: biomarker, contrasts, estimate, p.value
ebfit  <- runEbfit(tb)
BigTab <- runBigTab(ebfit, adj = "BH", mcut = 1.3, pcut = 0.05)
# -> columns lgFCH_*, FCH_*, pvals_*, fdrs_*, StatusFCH1.3FDR0.05_*
```

Already have a DESeq2 / limma table? You can build the equivalent columns
yourself, or skip the BigTab and pass parallel vectors directly to the
enrichment pipeline (shown later).

## Summarising the differential expression

`process_degs()` counts up/down genes per contrast from the `Status_*` columns
and writes one Excel workbook (a `Summary` sheet plus one sheet of DEGs per
contrast):

```{r process-degs}
res <- process_degs(
  BigTab      = BigTab,
  cont.list   = c("Drug.W2vsBL", "Drug.W16vsBL"),
  FCH         = 1.3,
  pcut        = 0.05,
  method      = "fdr",
  output_xlsx = "DEG_summary.xlsx"
)
res$summary          # per-contrast Up / Down counts
res$degs$Drug.W16vsBL  # DEG symbols for one contrast
```

# Annotated heatmaps

BioRosa provides two heatmap functions. Both draw a row-Z-scored expression body
on the left and a per-contrast fold-change table on the right, read from the
same BigTab; they differ only in how that table looks. Both **write a PDF** to
`grafname` (they do not return an inline figure).

| Function | Side table |
|---|---|
| `doAnnotatedHeatmap()` | One text column per contrast; numbers decimal-aligned with significance stars. |
| `doAnnotatedHeatmapGrid()` | A clean grid with **boxed, colour-filled group bands** parsed from the contrast names. |

## Grid variant — `doAnnotatedHeatmapGrid()`

The grid table groups contrasts into boxed, colour-filled header bands; cells
carry the signed fold change + stars, decimal-aligned per column.

```{r heatmap-grid}
doAnnotatedHeatmapGrid(
  grafname   = "heatmap_grid.pdf",
  mat        = expr_means,            # features x sample/group-mean columns
  annot.col  = annotx,                # row names define the column order
  colsplitx  = annotx$response,       # split the body into column groups
  BigTab     = BigTab,                # FCH_* / pvals_* columns
  BigTabcols = c("R.Drug.W2vsBL", "NR.Drug.W2vsBL", "RvsNR.Drug.W2vsBL"),
  setname    = "Immune proteins")
```

It auto-sizes the canvas and fonts to the gene-set size, parses the group bands
from the contrast names (here `R` / `NR` / `RvsNR`), and drops any dot-token that
is constant across contrasts from the sub-labels. Everything is overridable —
see `?doAnnotatedHeatmapGrid`.

## Side-table variant — `doAnnotatedHeatmap()`

The classic variant takes the log fold changes (`cfx`) and p-values (`fdx`) as
matrices (genes x contrasts) and appends one text column per contrast:

```{r heatmap-side}
contrasts <- c("R.Drug.W2vsBL", "NR.Drug.W2vsBL", "RvsNR.Drug.W2vsBL")
genes     <- rownames(BigTab)

cfx <- as.matrix(BigTab[genes, paste0("lgFCH_", contrasts)])  # log2 fold change
fdx <- as.matrix(BigTab[genes, paste0("pvals_", contrasts)])  # p-values

doAnnotatedHeatmap(
  grafname  = "heatmap_side.pdf",
  matx      = expr_means[genes, ],
  annot.col = annotx,
  colsplitx = annotx$response,
  cfx       = cfx,
  fdx       = fdx,
  setname   = "Immune proteins")
```

See `?doAnnotatedHeatmap`. Use the **grid** variant when you want grouped,
colour-banded contrasts for a slide or figure; the **side-table** variant when
you just need the per-contrast numbers beside the heatmap.

# Clinical-score trajectories

Alongside gene expression, BioRosa plots **longitudinal clinical scores**
(EASI, IGA, mSWAT, DLQI, …) from a **long-format** table — one row per
patient-visit with columns for the subject, the visit, the score, and the
treatment arm. Two functions cover the descriptive and the model-based views.

| Function | Shows |
|---|---|
| `doTrajectory()` | Observed data: per-patient "spaghetti" lines and/or a group-mean line with error bars. |
| `doTrajectoryModel()` | Mixed-model **estimated means** (or change / % change from baseline) with error bars and **significance stars**. |

Both accept `impute = c("none", "locf", "mice")` to fill missing visits
(last-observation-carried-forward, or predictive-mean-matching multiple
imputation) before plotting.

## Observed trajectories

```{r traj-observed}
# Per-patient spaghetti + bold group-mean line, faceted by arm
doTrajectory(clin, subject = "record_id", time = "Visit",
             score = "EASI", group = "Trt",
             time_levels = c("W0", "W4", "W8", "W16"))

# The classic mean +/- SE trajectory (no spaghetti)
doTrajectory(clin, subject = "record_id", time = "Visit",
             score = "EASI", group = "Trt",
             spaghetti = FALSE, error = "se")
```

## Mixed-model estimated means with stars

`doTrajectoryModel()` fits `score ~ group * time` with a per-subject random
intercept (`nlme::lme` by default, or `lme4::lmer`), then uses `emmeans` to get
the estimated means and the contrasts behind the stars:

* **within-arm** stars test each visit against `baseline` (drawn below each point);
* **between-arm** stars test the difference in change-from-baseline versus the
  `reference` arm at each visit (drawn above).

```{r traj-model}
res <- doTrajectoryModel(clin, subject = "record_id", time = "Visit",
                         score = "EASI", group = "Trt", baseline = "W0",
                         value = "change",        # or "mean", "pchange"
                         reference = "Pbo",
                         impute = "locf")
res$plot          # publication-ready ggplot
res$estimates     # per-arm/visit estimates, p-values, and star labels
```

## Reusing significance from a BigTab

The stars are computed from the fitted model by default, but you can **override**
them with p-values you already have — either a plain table (`time`, `p.value`,
optionally `group`) or a **BigTab** row, reusing the same `pvals_<contrast>`
columns that drive the heatmaps:

```{r traj-stars}
# Read pvals_Drug.W16vsBL, pvals_Pbo.W16vsBL, ... from a BigTab row
res <- doTrajectoryModel(clin, subject = "record_id", time = "Visit",
                         score = "EASI", group = "Trt", baseline = "W0",
                         value = "change",
                         pvals = BigTab, score_row = "EASI",
                         baseline_label = "BL")   # the token in ...vsBL
```

# One-call enrichment: `enrichment_onestep()`

`enrichment_onestep()` runs the full pathway-analysis battery in a single call
and writes every table and figure to `output_dir`. It is built around an
important distinction:

* **GSEA** (`fgsea` and `gseGO`) runs **once on the full ranked gene list** —
  direction is read from the sign of the NES. (Running GSEA on a pre-filtered or
  up/down-split list is statistically wrong.)
* **ORA** (`enrichGO` / `enrichKEGG` / `enrichPathway` / Hallmark, plus the
  OpenXGR-style RXGR engine) runs on the **UP and DOWN gene sets** — where the
  split is valid.

## From a BigTab + contrast

The most convenient entry point reads the log fold changes and p-values
straight from a BigTab for one contrast suffix:

```{r enrich-bigtab}
res <- enrichment_onestep(
  BigTab     = BigTab,
  contrast   = "Drug.W16vsBL",   # reads lgFCH_Drug.W16vsBL + pvals_Drug.W16vsBL
  output_dir = "results/Drug.W16vsBL"
)
```

## From parallel vectors

```{r enrich-vectors}
res <- enrichment_onestep(
  genes          = deg$SYMBOL,
  log2FoldChange = deg$log2FoldChange,
  padj           = deg$padj,
  output_dir     = "results/enrichment"
)
```

## Useful options

```{r enrich-options}
res <- enrichment_onestep(
  BigTab   = BigTab,
  contrast = "Drug.W16vsBL",

  # 1. Define the ORA UP/DOWN sets from the BigTab status column (1 = up,
  #    -1 = down) instead of recomputing from thresholds. On by default when
  #    a status column is present.
  ora_from_status = TRUE,
  status_prefix   = "Status",     # matches Status_... and StatusFCH1.3FDR0.05_...

  # 2. Choose the GSEA ranking statistic. "FCH" amplifies large fold changes
  #    in the weighted enrichment score; "signed_pval" = sign(lgFCH)*-log10(p).
  rank_by = "lgFCH",              # or "FCH", "signed_pval"

  # 3. Toggle the OpenXGR-style Fisher / odds-ratio engine (on by default).
  run_rxgr = TRUE,

  output_dir = "results/Drug.W16vsBL"
)
```

The returned list has four elements:

* `res$fgsea` — per-collection FGSEA (`Hallmark`, `GO_BP`, `Reactome`).
* `res$ora`   — `$UP` / `$DOWN`, each with `GO`, `Reactome`, `KEGG`, `Hallmark`.
* `res$gsea`  — a single `gseGO` result.
* `res$rxgr`  — `$UP` / `$DOWN`, each a per-collection RXGR result
  (`$table`, `$dotplot`, `$forest`).

…and `output_dir/` is populated with the matching Excel tables and
dotplot / treeplot / cnetplot / ridgeplot / forest figures.

# The OpenXGR-style enrichment engine (RXGR)

The Fisher's-exact engine that powers `res$rxgr` is also exposed directly. It
takes a query gene vector and any **named list of gene sets** — including the
bundled `guttman_pathways` — and reports the same statistics as the OpenXGR web
server: enrichment z-score, odds ratio with 95% CI, and FDR.

```{r osea, eval = TRUE, message = FALSE, fig.width = 6, fig.height = 4}
# A small, self-contained example (no Bioconductor / network needed).
sets  <- guttman_pathways[lengths(guttman_pathways) > 0]
query <- sets[["Immune Genes (EG)"]]

eset <- oSEA(query, sets, min.overlap = 3, size.range = c(5, 3000),
             verbose = FALSE)
head(oSEAextract(eset, sortBy = "or")[, c("name", "nO", "or", "adjp")])
```

The two companion figures — a dotplot (z-score vs significance) and a forest
plot (log2 odds ratio with 95% CIs) — are produced from the same object:

```{r osea-figs, eval = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5}
oSEAforest(eset, top = 10)
```

`rxgr_enrichment()` is the one-call wrapper that runs `oSEA()` and writes the
table + both figures to disk, mirroring the OpenXGR downloads.

# The bundled pathway list

`guttman_pathways` is the lab's curated gene-set collection (May 2026 release):
a named list of gene-symbol vectors, with many signatures split into directional
`... Up` / `... Down` halves.

```{r data, eval = TRUE}
length(guttman_pathways)                       # number of gene sets
head(names(guttman_pathways), 4)               # set names
length(guttman_pathways[["Immune Genes (EG)"]])  # genes in one set
```

A common pattern is to restrict a gene set to what you actually measured before
feeding it to a heatmap or scoring routine:

```{r data-subset}
present <- intersect(guttman_pathways[["Immune Genes (EG)"]], rownames(expr_mat))
```

# Where to go next

* `?BioRosa` — package overview and a map of every workflow.
* `?doAnnotatedHeatmap`, `?doAnnotatedHeatmapGrid` — the two annotated heatmaps.
* `?doTrajectory`, `?doTrajectoryModel` — clinical-score trajectory plots.
* `?enrichment_onestep` — the full enrichment pipeline and its options.
* `?oSEA`, `?oSEAforest`, `?oSEAdotplot` — the OpenXGR-style engine.
* `?runEbfit`, `?runBigTab`, `?process_degs` — building and summarising a BigTab.

```{r session, eval = TRUE}
sessionInfo()
```
